---
title: "Diamond sizes"
date: 2016-08-25
output: 
  html_document:
    df_print: paged
    
---

```{r setup, include = FALSE}
library(ggplot2)
library(dplyr)

smaller <- diamonds %>% 
  filter(carat <= 2.5)

knitr::opts_chunk$set(echo=F,comment = "",cache=T)
```

We have data about `r nrow(diamonds)` diamonds. Only 
`r nrow(diamonds) - nrow(smaller)` are larger than
2.5 carats. The distribution of the remainder is shown
below:

```{r, echo = FALSE}
smaller %>% 
  ggplot(aes(carat)) + 
  geom_freqpoly(binwidth = 0.01)
```

The most striking feature I guess would be all the spikes in count at "clean" fractions or integers of carat. 

# Exercise 27.4.7 #1 description

I'm exploring how diamonds vary by cut, colour and clarity. Diamond cuts are either:

```{r}
cat(levels(unique(diamonds$cut)))
```

Diamond colors are either:

```{r}
cat(levels(unique(diamonds$color)))
```

And diamond clarity are:

```{r}
library(ggplot2)
diamonds %>% 
  ggplot() +
  geom_bar(aes(clarity,..count..,fill=clarity)) +
  ggthemes::theme_tufte()
```

We can see below how the clarity of the diamond doesn't change very much in distribution when stratified by the cut of the diamond. 

```{r}
diamonds %>% 
  ggplot() +
  geom_bar(aes(clarity,..count..,fill=clarity)) +
  facet_wrap(~cut) +
  ggthemes::theme_tufte()
```

Same deal with the color of the diamond. 

```{r}
diamonds %>% 
  ggplot() +
  geom_bar(aes(clarity,..count..,fill=clarity)) +
  facet_wrap(~color) +
  ggthemes::theme_tufte()
```

# Exercise 27.4.7 #2 description

The top 20 diamond sizes are:

```{r}
tmp <- diamonds %>% 
  top_n(carat,n = 20) %>% 
  pull(carat) 

top20 <- unique(tmp)

cat(top20)
```

The clarity in these diamonds have only 3 clarities, where all but one have one clarity. 3.01 carat diamonds are either Sl2 or Il clarity. 

```{r}
top20_df <- diamonds %>% 
  filter(carat %in% top20) 

top20_df %>% 
  ggplot() +
  geom_bar(aes(clarity,..count..,fill=clarity)) +
  facet_wrap(~factor(carat))
```

Same with the color of the diamond. 

```{r}
top20_df %>% 
  ggplot() +
  geom_bar(aes(color,..count..,fill=color)) +
  facet_wrap(~factor(carat))
```

The price and carat are linearly correlated after about 3.2 carat diamonds. However, this may be due to small sample size. 

```{r}
top20_df %>% 
  ggplot(aes(carat,price)) +
  geom_point() +
  geom_smooth() +
  ggthemes::theme_tufte()
```

```{r, results="asis"}
top20_df
```





# Exercise 27.4.7 #4 description

Here's 4 chunks:

```{r a}
d <- readr::read_csv("diamonds.csv")
lubridate::now()
```

```{r b,dependson="a"}
d %>% 
  group_by(color) %>% 
  count(cut)
lubridate::now()
```

```{r c,dependson="a"}
d%>% 
  ggplot() +
  geom_bar(aes(clarity,..count..))
lubridate::now()
```

```{r d, dependson=c("c","b")}
d %>% 
  ggplot() +
  geom_point(aes(carat,price))
lubridate::now()
```




