---
name: "Nicholas Giangreco"
eracommons: "14654270"
position: "Quantitative Translational Scientist"
education:
  - degree: "BS"
    school: "University of Rochester"
    date: "5/2014"
    field: "Biochemistry"
  - degree: "PhD"
    school: "Columbia University"
    date: "11/2021"
    field: "Cellular, Molecular, and Biomedical Studies"
researchsupport:
  ongoing:
  completed:
    - grant: "R35GM131905"
      pi: "Nicholas Tatonetti"
      dates: "2019-2024"
      title: "Data-driven drug discovery: investigating the molecular mechanisms of safety and efficacy"
      description: "Prescription medicines are an essential component of modern medicine, however, while these medicines work well for some patients, they cause dangerous side effects in others. The lack of diversity in clinical trials means that these effects may disproportionately affect minorities and underrepresented patient populations. Using the electronic health records, I propose to investigate the reasons for adverse drug reactions in patients of different ages, sexes, genders, and ancestries."
      role: "Investigator"
output:
  biosketchr::nih_biosketch:
    latex_engine: xelatex
    keep_tex: yes
bibliography: yourbibfile.bib
bibliographystyle: apalike
---

```{r global_options, include=FALSE}
library(dplyr)
library(biosketchr)

## This saves you from having to do this for every chunk
knitr::opts_chunk$set(fig.path = 'figures/',
                      echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      results = "asis")

# load positions/honors data
d <- readxl::read_xlsx("positions-honors-nih.xlsx")

```

<!--
The structure of this skeleton should stay the same. Add your specifics :)
-->


# Personal Statement {-}

I have led and contributed to precision medicine research as a PhD trainee and currently in the pharmaceutical industry. My PhD thesis work led to multiple publications including a database of pediatric-specific adverse drug effect signals aligning with dynamic physiological processes during child development. I led the data science efforts to develop an interpretable and robust machine learning algorithm that led to hypothesized biological mechanism for a fatal, idiopathic graft dysfunction after heart transplant surgery.

```{r statement_citations}
# replace these keys with your own
make_numbered_citations("peddb", "pedreview", "kks", "klkb1")
```


# Positions and Honors {-}

## Positions and Employment {-}
```{r positions}
d |> 
  dplyr::filter(section == "positions") |> 
  biosketchr::make_datetbl_nih(start_date, end_date, description, order)
```

## Other Experience and Professional Memberships {-}

```{r other_exp}
d |> 
  dplyr::filter(section == "other_exp") |>  
  biosketchr::make_datetbl_nih(start_date, end_date, description, order)
```

## Honors {-}

```{r honors}
d |> 
  dplyr::filter(section == "honors") |> 
  biosketchr::make_datetbl_nih(start_date, end_date, description, order, FALSE)
```

# Contribution to Science {-}

\begin{enumerate}

\item Side effects are significant safety concerns in pediatric drug treatment but are rarely captured during clinical trials and are severely underreported post-market. Moreover, variations in metabolism and physiology as children grow and develop complicate detection of drug safety signals across child development. We developed a novel machine learning approach for identifying ontogenic-mediated adverse event mechanisms. We then made a database of 500,000 drug safety signals, called KidSIDES, freely available and browsable by a web application.

```{r contribution_citations1}
# replace these keys with your own
make_numbered_citations("peddb", "pedsim", "pedreview")
```


\item Primary graft dysfunction (PGD) is the leading cause of early mortality after heart transplant. Pre-transplant predictors of PGD remain elusive and its etiology remains unclear. A novel, patented machine learning algorithm identified pre-transplant level of KLKB1 is a robust predictor of post-transplant PGD. Our algorithm enabled the hypothesis that upregulation of coagulation cascade components of the kallikrein-kinin system (KKS) and downregulation of kininogen prior to transplant were associated with survival after transplant.

```{r contribution_citations2}
# replace these keys with your own
make_numbered_citations("kks", "klkb1")
```

\end{enumerate}

<!--
Can include additional contributions like the ones above, starting each with
\item, followed by a short description, follwed by a chunk using 
`make_numbered_citations`. Make sure these all appear before `end{enumerate}`
-->




